Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-06-26 doc/invt.tex
small text changes

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of flat-distribution sampling methods
  for free energy calculation including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the error in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme.
  %
  While the asymptotically optimal schedule for
  the single-bin updating scheme (commonly used in WL)
  is given by an inverse-time formula,
  that for the Gaussian updating scheme (commonly used in metadynamics)
  is often more complex.
  %and the initial updating magnitude is ideally half of
  %the previous equilibrium value.
  %
  Further,
  we show that the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  These bandpass updating schemes aim at
  partially flattening a few selected histogram modes
  and their optimal schedule
  is also given by the inverse-time prescription.
  %
  Several previous flat-distribution-sampling algorithms
  can be cast as special cases of the bandpass updating schemes.
  %
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry
that can provide insight into an array of phenomena not easily studied
with traditional experiments.
%
Given a system,
often the task is to compute
a distribution, $p(z)$,
along a collective variable, $z$, by sampling via either Monte Carlo\cite{
  frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The dimensionless free energy, or potential of mean force (PMF),
along the collective variable, $z$,
is $-\ln p(z)$.
%
However, accurately estimating the PMF is often difficult
when the collective variable is energetically restricted to local regions
of the complicated, multimodal free energy surface.
%
Thus,
to overcome these energetic barriers and
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF.
%
Ideally, then, sampling yields a flat distribution
along $z$\cite{mezei1987, berg1992, *lee1993,
wang2001, wang2001pre,
huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}.
%
The flat-distribution sampling can be used to accelerate
sampling of slow transitions that otherwise would not readily
be observed with traditional simulations.

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, wang2001pre}
and metadynamics\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, micheletti2004}
among others\cite{kim2006, *kim2007, junghans2014,
  langfeld2012, pellegrini2014}.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Often, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
We will refer to the two updating schemes
as the single-bin
and Gaussian updating schemes, respectively.
%
While the former is popular for MC simulations\cite{wang2001,
  wang2001pre, kim2006, *kim2007},
the latter is more popular for MD simulations
as it provides a smoother bias potential
that can be readily differentiated for force calculation.\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, junghans2014}
%
In both cases,
regular updates to the bias potential
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

@@

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian versions.
%
We will map the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
%Ideally,
%the optimal initial updating magnitude
%can be set to
%half of the previous equilibrium value,
%which amounts to a shift of the origin of time
%in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter, however,
allow one to partially flatten
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also an inverse-time schedule.
%
A previous FDS algorithm\cite{langfeld2012, pellegrini2014}
can be cast as a special case of the bandpass updating scheme,
and a similarly application can help other FDS
algorithms\cite{neuhaus2006, *neuhaus2007, zhu2012}
adjust the parameters of the bias potential automatically.

@@

%For a large system at temperatures of interest,
%the distribution, $p_i$, often tends to
%be localized,
%%
%and to explore the global shape
%of the PMF, $-\ln p_i$,
%it is often advantageous to perform
%biased sampling that targets
%a wider and smoother distribution, $\rho_i$.
%%
%%Here, we refer to simulations that target
%%a flat or nearly flat distribution
%%as entropic or multicanonical sampling.
%
%To do so, we introduce a bias potential, $u_i$,
The difficulty of sampling a multimodal distribution
with deep valley(s) can sometimes be eased by
sampling a smoother biased distribution, $\rho_i$.
%
%
%where the proportionality constant, $C_i$,
%is found from the normalization condition,
%$\sum_{i = 1}^n \pi_i = 1$,
%in the second step.
%
%In particular,
%to make $\pi_i$ flat,
%the bias potential $u_i$
%must coincide with $\ln p_i$,
%up to an additive constant.
%%
%In practice, $u_i$ may be non-optimal or contain error,
%and thus $\pi_i$ need not
%coincide with the intended distribution $\rho_i$.
%
We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $\rho_i$ (i.e. $\rho_i \equiv 1/n$ for $n$ bins)
or nearly flat one\cite{
dayal2004, *trebst2004, barducci2008, singh2011}.
%
By introducing a bias potential, $u_i$,
the biased equilibrium distribution becomes
%
\begin{equation}
  \pi_i
  %=
  %C_i \, p_i \, e^{-u_i}
  =
  \frac{                p_i \, e^{-u_i} }
       { \sum_{j = 1}^n p_j \, e^{-u_j} }
  \propto
  p_i \, e^{-u_i}
  ,
  \notag
  %\label{eq:pi_p_V}
\end{equation}
%
and our aim is to find a bias potential
such that the resulting $\pi_i$ would coincide
with the desired $\rho_i$.
%
Upon convergence, the bias potential would satisfy
%
\begin{equation}
  u_i \to \ln \frac{p_i}{\rho_i} + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
%according to Eq. \eqref{eq:pi_p_V}.
%which allows us to deduce the PMF, $-\ln p_i$,
%directly from the bias potential.
%
and the PMF, $-\ln p_i$, can be readily
deduced from the bias potential. % $u_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p_i$ is replaced by $\ln Z_i$,
  $\rho_i$ is the weight $\zeta_i$,
  and the bias potential, $u_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

@@

An other class of adaptive FDS methods,
including the WL algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002, *laio2008, *barducci2011, *sutto2012},
address the above shortcoming
by continuously and incrementally updating the bias potential,
and we will focus on these methods below.
%%
%%These methods incrementally update the bias potential
%%and can approximately sample the target distribution.
%%
%However,
%the on-the-fly updates break the microscopic reversibility
%of the underlying sampling mechanism.
%Further, failure to properly reduce the magnitude of
%the histogram-based increments
%would lead to roughness or errors in the final estimate of the PMF.
%%
%%Thus, for a conservative practitioner,
%%the E-FDS methods appear more appealing
%%once a sufficiently accurate bias potential is available.
%%
%To minimize this effect,
%many adaptive methods adopt a schedule to
%reduce the updating magnitude
%over the course of simulation\cite{
%marsili2006,
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%barducci2008}.
%%
%%Thus, in late stages of a long simulation,
%%the updating becomes negligible and
%%the adaptive simulation tends to
%%the corresponding equilibrium one.
%%
%%if the bias potential were converged to the exact value,
%%this adaptive FDS tends to the ideal equilibrium counterpart,
%%particular equilibrium simulation can be thought as
%%the optimal asymptotic limit of the adaptive counterpart,
%%
%For example, one can show that
%if the magnitude is controlled by
%the optimal inverse-time prescription\cite{
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%morozov2007, zhou2008,
%komura2012, *caparica2012, *caparica2014},
%the WL algorithm is equivalent to the ideal equilibrium counterpart
%in improving the precision of the bias potential,
%thus allowing for maximum efficiency
%(cf. Appendix \ref{sec:equilerr}).
%%the bias potential in the WL algorithm
%%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%%with the precision of the histogram as simulation lengthens.
%
%
%\subsubsection{Updating schemes}
%
%
%
%We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $u_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { \rho_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
%including the Gaussian updating scheme used in metadynamics,
the updating also involves neighboring bins:
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { \rho_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
The Gaussian updating scheme used in metadynamics
is an example, with $w_{ij}$ being a Gaussian centered on $j$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

While adaptive FDS simulations
can often sample the desired distribution,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors that need to be
removed by gradually decreasing the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%One can show that under constant magnitude,
%$\alpha(t) = \alpha > 0$,
%the distribution collected from
%an adaptive FDS simulation
%is identical to $\rho_i$ in the long time or asymptotic limit.
%
%However, as the above updates
%change $u_i(t)$ continuously,
%the equilibrium distribution,
%Eq. \eqref{eq:pi_p_V}, no longer holds
%at all times,
%%
%The source of the deviation is two-fold.
%%
%First, since $u_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%

@@

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of updating steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
This can be derived from a general result\cite{
  robbins1951, pellegrini2014}
and an intuitive explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight,
the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
%We may also derive it
%from optimally planning the simulation lengths in the stages
%and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

@@

We define the net error in the bias potential
by deducting two contributions.
%
First, to deduct the target value given by
Eq. \eqref{eq:Vi_target},
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  u_i(t)
  -
  \ln \frac { p_i }
            { \rho_i }
  ,
  \notag
  %\label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i(t)$ are
equivalent to the updates to $u_i(t)$,
since the shift
remains a constant during the course of simulation.
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                \rho_i \, e^{-v_i} }
%       { \sum_{j = 1}^n \rho_j \, e^{-v_j} }
%  \propto
%  \rho_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

@@

We can split the histogram into
a deterministic averaged part, $\langle f_i(t) \rangle$,
and a random fluctuating part, $\zeta_i(t)$,
%
\begin{equation}
  f_i(t) =
  \langle f_i(t) \rangle
  +
  \zeta_i(t)
  .
  \notag
  %\label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consisting of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle f_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { \rho_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { \rho_i }
  =
  \frac{                          e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n \rho_j \, e^{-v_{*j}(t)} }
  %\notag
  %\\
  %&
  \approx
  1 - v_{*i}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { \rho_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n \rho_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n \rho_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average and using Eq. \eqref{eq:fav1}, we get
%
\begin{align}
  \langle f_{*i}(t) \rangle
  \equiv
  \langle f_i(t) \rangle - 1
  &\approx
  - v_{*i}(t)
  .
  \label{eq:sh_ave}
\end{align}

We will approximate the histogram fluctuation part, $\zeta_i(t)$,
hence its deviation from the bin average, $\zeta_{*i}(t)$,
as white noise such that
\begin{equation}
  \langle \zeta_{*i}(t) \, \zeta_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \notag
  %\label{eq:G_def}
\end{equation}
for some $G_i$,
and $\delta(t)$ is the Dirac delta function.
%
We expect the approximation to hold for simulation
much longer than the autocorrelation time.
%
%For non-white noise,
%the equivalent value of $G_i$ can be approximately obtained as
%$\sum_{t = -\infty}^\infty \langle \zeta_{*i}(t) \, \zeta_{*i}(0) \rangle$.

From Eqs.
\eqref{eq:vt_diffeq}-\eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  =
  \alpha(t) \, \left[ \langle f_{*i}(t) \rangle + \zeta_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \zeta_{*i}(t) \right]
  ,
  \notag
  %\label{eq:dxdt_singlebin}
\end{equation}
%
and the formal solution is
%
\begin{equation}
  v_{*i}(T)
=
  v_{*i}(0) \, e^{-q(T)}
+
\int_0^T
  \dot \vartheta\bigl( q(t) \bigr) \, \zeta_{*i}(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
\vartheta(q')
&\equiv
e^{q' - q(T)}.
\label{eq:theta_def}
\end{align}

@@

\begin{figure}[h]
\begin{center}
  \begin{tikzpicture}[thick, scale=1.25, every node/.style={scale=0.8}]
    \node [] at (-0.3, 2.0) {(a)};
    % distribution
    \draw[domain=0:2.5, variable=\x, dashed, gray, very thick]
      plot({\x}, {2*exp(-\x)});

    % shade
    \fill [gray!30!white, domain=0.7:2.2, variable=\x]
      (0.7, 0) --
      plot ({\x}, {2*exp(-\x)})
      -- (2.2, 0) -- cycle;

>>>>

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of flat-distribution sampling methods
  for free energy calculation including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the error in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme.
  %
  While the asymptotically optimal schedule for
  the single-bin updating scheme (commonly used in WL)
  is given by an inverse-time formula,
  that for the Gaussian updating scheme (commonly used in metadynamics)
  is often more complex.
  %and the initial updating magnitude is ideally half of
  %the previous equilibrium value.
  %
  Further,
  we show that the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  These bandpass updating schemes aim at
  partially flattening a few selected histogram modes
  and their optimal schedule
  is also given by the inverse-time prescription.
  %
  Constructed from orthogonal polynomials,
  bandpass updating schemes serve as generalizations
  of the Langfeld-Lucini-Rago algorithm
  and can be used for parameter tuning for the method of Gaussian ensemble
  and umbrella sampling.
  %
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry
that can provide insight into an array of phenomena not easily studied
with traditional experiments.
%
Given a system,
often the task is to compute
a distribution, $p(z)$,
along a collective variable, $z$, by sampling via either Monte Carlo\cite{
  frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The dimensionless free energy, or potential of mean force (PMF),
along the collective variable, $z$,
is $-\ln p(z)$.
%
However, accurately estimating the PMF is often difficult
when the collective variable is energetically restricted to local regions
of the complicated, multimodal free energy surface.
%
Thus,
to overcome these energetic barriers and
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF.
%
Ideally, then, sampling yields a flat distribution
along $z$\cite{mezei1987, berg1992, *lee1993,
wang2001, wang2001pre,
huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}.
%
The flat-distribution sampling can be used to accelerate
sampling of slow transitions that otherwise would not readily
be observed in traditional simulations.

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, wang2001pre}
and metadynamics\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, micheletti2004}
among others\cite{kim2006, *kim2007, kim2010, junghans2014,
  langfeld2012, pellegrini2014,
  maragliano2006, *abrams2008,
  zheng2010}.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Often, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.\cite{junghans2014}
%
We will refer to the two updating schemes
as the single-bin
and Gaussian updating schemes, respectively.
%
While the former is popular for MC simulations\cite{wang2001,
  wang2001pre, kim2006, *kim2007},
the latter is more popular for MD simulations
as it provides a smoother bias potential
that can be readily differentiated for force calculation.\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, junghans2014}
%
In both cases,
regular updates to the bias potential
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one has to gradually decrease
the updating magnitude of the bias.

@@

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian versions.
%
We will map the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
%Ideally,
%the optimal initial updating magnitude
%can be set to
%half of the previous equilibrium value,
%which amounts to a shift of the origin of time
%in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter allow one to partially flatten
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also an inverse-time schedule.
%
Constructed from orthogonal polynomials,
bandpass updating schemes serve as
generalizations of the Langfeld-Lucini-Rago (LLR)
algorithm\cite{langfeld2012, pellegrini2014},
and can help other FDS
algorithms\cite{neuhaus2006, *neuhaus2007, zhu2012}
adjust the parameters of the bias potential automatically.

@@

The difficulty of sampling a multimodal distribution
with deep valley(s) can sometimes be eased by
sampling a smoother biased distribution, $\rho_i$.
%
%
We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $\rho_i$ (i.e. $\rho_i \equiv 1/n$ for $n$ bins)
or a nearly flat one\cite{
dayal2004, *trebst2004, barducci2008, singh2011}.
%
The biased equilibrium distribution
under a bias potential, $u_i$, is given by
%
\begin{equation}
  \pi_i
  %=
  %C_i \, p_i \, e^{-u_i}
  =
  \frac{                p_i \, e^{-u_i} }
       { \sum_{j = 1}^n p_j \, e^{-u_j} }
  \propto
  p_i \, e^{-u_i}
  ,
  \notag
  %\label{eq:pi_p_V}
\end{equation}
%
and our aim is to find a bias potential
such that the resulting $\pi_i$ would coincide
with the desired $\rho_i$.
%
Upon convergence, the bias potential would satisfy
%
\begin{equation}
  u_i \to \ln \frac{p_i}{\rho_i} + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
%according to Eq. \eqref{eq:pi_p_V}.
%which allows us to deduce the PMF, $-\ln p_i$,
%directly from the bias potential.
%
and the PMF, $-\ln p_i$, can be readily
deduced from the bias potential. % $u_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p_i$ is replaced by $\ln Z_i$,
  $\rho_i$ is the weight $\zeta_i$,
  and the bias potential, $u_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

@@

Below we will focus on another class of adaptive FDS methods,
including the WL algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002, *laio2008, *barducci2011, *sutto2012},
address the above shortcoming
by continuously and incrementally updating the bias potential.
%
%\subsubsection{Updating schemes}
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $u_i(t)$, is updated
at each step $t$ as
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { \rho_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
%including the Gaussian updating scheme used in metadynamics,
the updating also involves neighboring bins:
%
\begin{equation}
  u_i(t+1)
  =
  u_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { \rho_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
The Gaussian updating scheme used in metadynamics
is an example, with $w_{ij}$ being a Gaussian centered on $j$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

While adaptive FDS simulations
can often sample the desired distribution,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors that need to be
reduced by gradually decreasing the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.

@@

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of updating steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
This general result\cite{
  robbins1951, pellegrini2014}
has an intuitive explanation\cite{
  marsili2006, barducci2008}
that interprets the bias potential
as a runtime average computed from all data points collected so far,
which absorbs the contribution of the newest data point
by its due weight,
the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
%We may also derive it
%from optimally planning the simulation lengths in the stages
%and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

@@

We define the net error in the bias potential
by deducting two contributions.
%
First, to deduct the target value given by
Eq. \eqref{eq:Vi_target},
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  u_i(t)
  -
  \ln \frac { p_i }
            { \rho_i }
  ,
  \notag
  %\label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i(t)$ are
equivalent to the updates to $u_i(t)$,
since the shift
remains a constant during simulation.
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                \rho_i \, e^{-v_i} }
%       { \sum_{j = 1}^n \rho_j \, e^{-v_j} }
%  \propto
%  \rho_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

@@

We can split the histogram into
a deterministic averaged part, $\langle f_i(t) \rangle$,
and a random fluctuating part, $\zeta_i(t)$,
%
\begin{equation}
  f_i(t) =
  \langle f_i(t) \rangle
  +
  \zeta_i(t)
  .
  \notag
  %\label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consisting of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle f_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { \rho_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { \rho_i }
  =
  \frac{                          e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n \rho_j \, e^{-v_{*j}(t)} }
  %\notag
  %\\
  %&
  \approx
  1 - v_{*i}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { \rho_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n \rho_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n \rho_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average and using Eq. \eqref{eq:fav1}, we get
$\langle f_{*i}(t) \rangle \equiv \langle f_i(t) \rangle - 1 \approx - v_{*i}(t)$,
and
%
\begin{align}
  f_{*i}(t)
  &\approx
  - v_{*i}(t)
  +
  \zeta_i(t)
  .
  \label{eq:sh_ave}
\end{align}

We will approximate the histogram fluctuation part, $\zeta_i(t)$,
hence its deviation from the bin average, $\zeta_{*i}(t)$,
as white noise such that
\begin{equation}
  \langle \zeta_{*i}(t) \, \zeta_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \notag
  %\label{eq:G_def}
\end{equation}
for some $G_i$,
and $\delta(t)$ is the Dirac delta function.
%
We expect the approximation to hold for simulation
much longer than the autocorrelation time.

From Eqs.
\eqref{eq:vt_diffeq} and \eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  %=
  %\alpha(t) \, \left[ \langle f_{*i}(t) \rangle + \zeta_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \zeta_{*i}(t) \right]
  ,
  \notag
  %\label{eq:dxdt_singlebin}
\end{equation}
%
and the formal solution is
%
\begin{equation}
  v_{*i}(T)
=
  v_{*i}(0) \, e^{-q(T)}
+
\int_0^T
  \dot \vartheta\bigl( q(t) \bigr) \, \zeta_{*i}(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
\vartheta(q')
&\equiv
e^{q' - q(T)}.
\label{eq:theta_def}
\end{align}

@@

\begin{figure}[h]
\begin{center}
  \begin{tikzpicture}[thick, scale=1.4, every node/.style={scale=0.9}]
    \node [] at (-0.3, 2.0) {(a)};

    % distribution
    \draw[domain=0:2.0, variable=\x, dashed, gray, very thick]
      plot({\x}, {2*exp(-\x)});

    % shade
    \fill [gray!30!white, domain=0.7:1.609, variable=\x]
      (0.7, 0) --
      plot ({\x}, {2*exp(-\x)})
      -- (1.609, 0) -- cycle;

    % thick curve
    \draw[domain=0:1.609, variable=\x, very thick]
      plot({\x}, {2*exp(-\x)});
